---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# netmem: Social Network Measures using Matrices

<!-- badges: start -->
<!-- badges: end -->

The goal of `netmem` is to make available different measures to describe social networks using matrices.

- Author/mantainer: [Alejandro Espinosa-Rada](https://www.research.manchester.ac.uk/portal/en/researchers/alejandro-espinosa(4ed72800-e02b-47a8-a958-640b6a07f563).html) 

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("anespinosa/netmem")
```

## Example

```{r warning=FALSE}
library(netmem)
```

Let's assume that we have a multilevel network with three adjacency matrices and two incident matrices between them.

- `A1`: Adjacency Matrix of the level 1
- `B1`: Incident Matrix between level 1 and level 2
- `A2`: Adjacency Matrix of the level 2
- `B2`: Incident Matrix between level 2 and level 3
- `A3`: Adjacency Matrix of the leve 3

Create the data
```{r}
A1 <- matrix(c(0,1,0,0,1,
              1,0,0,1,1,
              0,0,0,1,1,
              0,1,1,0,1,
              1,1,1,1,0), byrow=TRUE, ncol=5)
              
B1 <- matrix(c(1,0,0,
              1,1,0,
              0,1,0,
              0,1,0,
              0,1,1), byrow=TRUE, ncol=3)

A2 <- matrix(c(0,1,1,
              1,0,0,
              1,0,0), byrow=TRUE, nrow=3)

B2 <- matrix(c(1,1,0,0,
              0,0,1,0,
              0,0,1,1), byrow=TRUE, ncol=4)

A3 <- matrix(c(0,1,1,1,
              1,0,0,0,
              1,0,0,1,
              1,0,1,0), byrow=TRUE, ncol=4)

B3 <- matrix(c(1,0,0,0,0,
              0,1,0,1,0,
              0,0,0,0,0,
              0,0,0,0,0), byrow=TRUE, ncol=5)

```

Explore the degree centrality of an adjacency matrix and a bipartite matrix
```{r results='hide'}
# Degree centrality of the first matrix
gen_degree(A1, digraph = FALSE)

# Degree centrality of one of the incident matrices
gen_degree(B1, bipartite = TRUE)

```

We can also check the triad census assuming **U|MAN**
```{r results='hide'}
triad_uman(A1)
```

Also, we would calculate the level of constraint of one particular actor (e.g. actor `5`) 
```{r results='hide'}
eb_constraint(A1, ego=5)
```

Now, we can calculate the degree centrality of the entire stucture
```{r results='hide'}
multilevel_degree(A1, B1, A2, B2, A3, B3, complete = TRUE)
```
